---
title: "Assignment 1"
output: html_document
---

Objective: Determine whether daily concentrations of PM2.5 have decreased in CA over the last 15 years (2004-2019)

1. Read in the data using data.table(). For each of the two data sets, check the dimensions, headers, footers, variable names and variable types. Check for any data issues, particularly in the key variable we are analyzing. Summarize findings.
```{r}
library(data.table)
library(tidyverse)

PM_2004 <- data.table::fread("ad_viz_plotval_data_2004.xls")
#check dimensions 
dim(PM_2004)
nrow(PM_2004)
ncol(PM_2004)
#check top and bottom of data set
head(PM_2004)
tail(PM_2004)
#check variables 
str(PM_2004)
#check key variable
summary(PM_2004$`Daily Mean PM2.5 Concentration`)
mean(is.na(PM_2004$`Daily Mean PM2.5 Concentration`))
```
No missing values

```{r}
PM_2019 <- data.table::fread("ad_viz_plotval_data_2019.xls")
#check dimensions 
dim(PM_2019)
nrow(PM_2019)
ncol(PM_2019)
#check top and bottom of data set
head(PM_2019)
tail(PM_2019)
#check variables 
str(PM_2019)
#check key variable
summary(PM_2019$`Daily Mean PM2.5 Concentration`)
mean(is.na(PM_2019$`Daily Mean PM2.5 Concentration`))
```
No missing values.
