---
title: "Assignment 1"
output:
  - github_document
  - html_document
always_allow_html: true
---

Objective: Determine whether daily concentrations of PM2.5 have decreased in CA over the last 15 years (2004-2019)

1. Read in the data using data.table(). For each of the two data sets, check the dimensions, headers, footers, variable names and variable types. Check for any data issues, particularly in the key variable we are analyzing. Summarize findings.
```{r}
library(data.table)
library(tidyverse)

PM_2004 <- data.table::fread("ad_viz_plotval_data_2004.xls")
#check dimensions 
dim(PM_2004)
nrow(PM_2004)
ncol(PM_2004)
#check top and bottom of data set
head(PM_2004)
tail(PM_2004)
#check variables 
str(PM_2004)
#check key variable
summary(PM_2004$`Daily Mean PM2.5 Concentration`)
mean(is.na(PM_2004$`Daily Mean PM2.5 Concentration`))
```
No missing values

```{r}
PM_2019 <- data.table::fread("ad_viz_plotval_data_2019.xls")
#check dimensions 
dim(PM_2019)
nrow(PM_2019)
ncol(PM_2019)
#check top and bottom of data set
head(PM_2019)
tail(PM_2019)
#check variables 
str(PM_2019)
#check key variable
summary(PM_2019$`Daily Mean PM2.5 Concentration`)
mean(is.na(PM_2019$`Daily Mean PM2.5 Concentration`))
```
No missing values.

2. Combine the two years of data into one data frame. Use the Date variable to create a new column for year, which will serve as an identifier. Change the names of the key variables so that they are easier to refer in your code.
```{r}
#combine data 
PM_combined <- merge(x = PM_2004, y = PM_2019, all = TRUE)
head(PM_combined)
#check number of rows
PM_2004 %>% nrow()
PM_2019 %>% nrow()
PM_combined %>% nrow()
#create new column for year
PM_combined[, Year := format(as.Date(Date, format = "%m/%d/%Y"), "%Y")]
head(PM_combined)
```
3. Create a basic map in leaflet that shows the locations of the sites (make sure to use different colors for each year). Summarize spatial distribution of monitoring sites.
```{r}
library(leaflet)
leaflet() %>%
  addProviderTiles('CartoDB.Positron') %>%
   addCircles(
    data = PM_2019,
    lat=~SITE_LATITUDE, lng=~SITE_LONGITUDE,
    opacity = 1, fillOpacity = 1, radius = 10, color = "blue") %>%
   addCircles(
    data = PM_2004,
    lat=~SITE_LATITUDE, lng=~SITE_LONGITUDE,
    opacity = 1, fillOpacity = 1, radius = 10, color = "red"
    )
```

4. Check for any missing or implausible values of PM25 in the combined dataset. Explore the proportions of each and provide a summary of any temporal patterns you see in these observations.
```{r}
#check dimensions 
dim(PM_combined)
nrow(PM_combined)
ncol(PM_combined)
#check top and bottom of data set
head(PM_combined)
tail(PM_combined)
#check variables 
str(PM_combined)
#check key variable
summary(PM_combined$`Daily Mean PM2.5 Concentration`)
summary(PM_combined$`DAILY_AQI_VALUE`)
summary(PM_combined$`DAILY_OBS_COUNT`)
#check Na's
mean(is.na(PM_combined$`Daily Mean PM2.5 Concentration`))
mean(is.na(PM_combined$`Year`))
mean(is.na(PM_combined$`Date`))
mean(is.na(PM_combined$`Site ID`))
mean(is.na(PM_combined$`POC`))
mean(is.na(PM_combined$`DAILY_AQI_VALUE`))
mean(is.na(PM_combined$`Site Name`))
mean(is.na(PM_combined$`DAILY_OBS_COUNT`))
mean(is.na(PM_combined$`AQS_PARAMETER_CODE`))
#Remove implausible values
PM_combined_filtered <- PM_combined[`Daily Mean PM2.5 Concentration` >= 0,]
summary(PM_combined_filtered$`Daily Mean PM2.5 Concentration`)
```

5. Explore the main question of interest at three different spatial levels. Create exploratory plots (e.g. boxplots, histograms, line plots) and summary statistics that best suit each level of data (state, county, LA). Be sure to write up explanations of what you observe in these data.
```{r}
library(ggplot2)

#State 

ggplot(PM_combined_filtered) +
  geom_histogram(mapping = aes(x = `Daily Mean PM2.5 Concentration`)) + xlim(c(0,75))

 ggplot(data = PM_combined_filtered) +
  geom_point(mapping = aes(x = Year, y = `Daily Mean PM2.5 Concentration`, color = Year), position = "jitter")
 
box_plot_crop <- ggplot(data= PM_combined_filtered, aes(x= Year, y=`Daily Mean PM2.5 Concentration`, fill=Year))
box_plot_crop + geom_boxplot(outlier.shape = NA) + coord_cartesian(ylim=c(0,35))

PM_combined_filtered_2004 <- PM_combined_filtered[Year %like% "2004",]
PM_combined_filtered_2019 <- PM_combined_filtered[Year %like% "2019",]
summary(PM_combined_filtered_2004$`Daily Mean PM2.5 Concentration`)
summary(PM_combined_filtered_2019$`Daily Mean PM2.5 Concentration`)

#County 

PM_combined_filtered_county <- PM_combined_filtered[CBSA_NAME %like% "Los Angeles-Long Beach-Anaheim, CA",]
box_plot_crop_county <- ggplot(data= PM_combined_filtered_county, aes(x= Year, y=`Daily Mean PM2.5 Concentration`, fill=Year))
box_plot_crop_county + geom_boxplot(outlier.shape = NA) + coord_cartesian(ylim=c(0,35))

PM_combined_filtered_county_2004 <- PM_combined_filtered_county[Year %like% "2004",]
PM_combined_filtered_county_2019 <- PM_combined_filtered_county[Year %like% "2019",]
summary(PM_combined_filtered_county_2004$`Daily Mean PM2.5 Concentration`)
summary(PM_combined_filtered_county_2019$`Daily Mean PM2.5 Concentration`)

#Los Angeles

PM_combined_filtered_city <- PM_combined_filtered[`Site Name` %like% "Los Angeles-North Main Street",]
box_plot_crop_city <- ggplot(data= PM_combined_filtered_city, aes(x= Year, y=`Daily Mean PM2.5 Concentration`, fill=Year))
box_plot_crop_city + geom_boxplot(outlier.shape = NA) + coord_cartesian(ylim=c(0,45))

PM_combined_filtered_city_2004 <- PM_combined_filtered_city[Year %like% "2004",]
PM_combined_filtered_city_2019 <- PM_combined_filtered_city[Year %like% "2019",]
summary(PM_combined_filtered_city_2004$`Daily Mean PM2.5 Concentration`)
summary(PM_combined_filtered_city_2019$`Daily Mean PM2.5 Concentration`)

```

